Pool Model - A unified Mathematical\\ Modelling Framework for Food Science

### 2025-10-31 paper/pool_model.tex
tex: correct mistakes, delete unnecessary commands

\setlength{\tabcolsep}{1.3pt}
\setlength{\columnsep}{0.4cm}
\usepackage{acronym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{array}
\usepackage{bm}
\usepackage{booktabs}
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off, font=small]{caption}
\usepackage{diagbox}
\usepackage{float}
\usepackage[T1]{fontenc}
% \usepackage[a4paper,inner=2cm,outer=2cm,top=2.5cm,bottom=2.5cm]{geometry}
\usepackage{graphicx}% Include figure files
\usepackage{here}
\usepackage{hyperref}
\usepackage{lipsum}
\usepackage{listings}
\usepackage{listings-rust}
\usepackage{mathtools}
\usepackage{mathrsfs}
\usepackage{multirow}
\usepackage{newfloat}
\usepackage{onimage}
\usepackage{siunitx}
\usepackage{subcaption}
\usepackage{tabularx}
\usepackage{tikz}
\usepackage[dvipsnames]{xcolor}
\usepackage{xpatch}

@@

\newcommand{\beginsupplement}{%
  \renewcommand{\thesection}{}
  \setcounter{subsection}{0}
  \renewcommand{\thesubsection}{S\arabic{subsection}}
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}%
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}%
  \setcounter{listing}{0}
  \renewcommand{\thelisting}{S\arabic{listing}}
}

\newcommand{\mbP}{\textbf{P: }}
\newcommand{\mbC}{\textbf{C: }}
\newcommand{\mbx}{\mathbf{x}}
\newcommand{\mbm}{\mathbf{m}}
\newcommand{\mbp}{\mathbf{p}}
\newcommand{\mbu}{\mathbf{u}}
\newcommand{\mbF}{\mathbf{F}}
\newcommand{\mbH}{\mathbf{H}}
\newcommand{\cmtb}[1]{\textcolor{blue} {#1}}
\newcommand{\cmt}[1]{\textcolor{red} {[#1]}}
\newcommand{\etal}{{\textit{et al. }}}

@@

\cortext[cor1]{Corresponding Author}
\fntext[fn1]{jonas.pleyer.org}
\affiliation[1]{
    organization={Freiburg Center for Data Analysis and Modeling},
    addressline={Ernst-Zermelo Str. 1},
    city={Freiburg im Breisgau},
    citysep={},
    postcode={79104},
    country={Germany},
}

@@

\begin{abstract}
    Bacterial consortia are complex systems which are often comprised of multiple distinct species.
    Their dynamics on a population-based level often result from interactions that lead to
    inter-species cooperation or competition.
    Various empirical and inactivation models have been conceived and applied to problems related to
    food microbiology.
    However, the utilized models live on the population level, omitting any spatial effects such as
    heterogeneity or diffusion.
    In this work we present a novel generalized framework, called "pool model" which uses multiple
    distinct pools that can hold bacterial states of distinct species or extracellular nutrient
    abundance.
    It allows us to describe these systems mechanistically as variations of an underlying, more
    fundamental model.
    To illustrate the usefulness of the model, multiple different examples are reported.
    Finally, we construct an \acl{abm} for a given pool model, showcasing that the resulting model
    arises naturally from the individual behaviour of many cellular agents.
    We further apply this \ac{abm} to study the role of spatial effects on the dynamcis of bacterial
    growth.
\end{abstract}

@@

To determine the system of equations which describe the systems dynamics, we consider the flows between the different states.

@@

As a result, species $A$ was able to suppress species $B$, consuminl the larger amount of nutrients and reaching a larger bacterial count at the end.

@@

One of the examples of such a mechanisms is the production of antimicrobial compounds, inhibitors or toxins~\cite{wloch-salamon_effect_2008, chao_structured_1981}.

@@

This mutual activation benefits all involved species and can by easily included in our mathematical framework.

@@

nutritients~\cite{htuson_bacteriasurface_2013}.

@@

\textit{B. subtilis} which are capable of chemotactic motion of driven by oxygen

@@

The bacterial social interactions such as releas of toxins or public goods can also

@@

% ------------------------------------------------------------------------------------------------
\subsection{Constructing an Agent-Based Pool Model}
The following sections will demonstrate how a given \ac{ode} system can be translated into a
custom-built \ac{abm} which has only a few more overall parameters.
This procedure is a blueprint which can be followed to construct \acp{abm} for similar
\ac{ode}-based systems.\\
To construct our \ac{abm}, we use \texttt{cellular\_raza}~\cite{Pleyer2025}.
It allows us to define our bacterial agents from a clean slate, without having to rely and inherit
existing work.
With this flexibility, researchers are able to construct minimal models using only as many parameters which are necessary.
This ability, to precisely define and control the exact behavior of our cell-agents, enables the following parts.

@@

Additionally, we require our external components to be freely diffusable with constants $D_R,D_I$.

@@

In subfigure (c), due to the heterogenous cell distribution (see also~\ref{fig:supplement-spatial-initial-config-cells}), overall inhibition of species $B$ is significantly reduced.

@@

This is expected, since processes such as diffusion and cells growing outward introduce another time component which is simply assumed to be instantanous in the \ac{ode} model.

@@

        homogenous (Hom.), heterogenous (Het.) or randomized (Rand.) initial configurations.

@@

The pool model presented here was used to describe the dynamics of 6 systems containing established models of bacterial growth.
Comparing our approach with the Baranyi-Roberts model which is widely used in food science, it is possible to obtain results which provide growth curves of very similar shape.
However, their analytical form is not completely identical.
The inclusion of additional metabolic processes or larger number of cell-species can be achieved by extending the list of possible states and defining transitions between them.
In addition we have shown how to treat interspecies interactions in the form of cooperation and competition.
By constructing an \ac{abm}, we have exemplified how close the devised model is to the individual biological processes that occur within the bacteria.
This mechanistic single-cell level of understanding of our \ac{abm} is a perspective which can not be easily replicated by other spatial models such as \acp{pde}.
We have investigated the effect of spatial effects in the form of varying diffusion parameters and different initial spatial distributions of bacteria.
Our findings reinforce that the \ac{ode} model is a good simplification in the highly-diffusive regime within a mostly homogenous spatial bacterial distribution.
However, deviations to these assumptions can lead to non-negligible effects, influencing the growth of the colony.

@@

We introduced the pool model as a novel framework to unify existing approaches and allows researchers to follow a mechanistic approach during model development.
It is able to capture the effects of many established models and provides a lens which is useful to generalize existing systems and results in better overall explainability.
By modeling effects such as mutual inhibition on a single-cell level within an \ac{abm}, we have exemplified that the pool model directly builds on these more fundamental concepts.
It can thus aid in the mechanistic understanding of the underlying processes, allowing researchers to deepen their konwledge about microbial growth within food science.
Going forward, this framework can provide a starting point for a more abstract mathematical formulation which should be able to cover a much wider class of systems.

@@

This means that every contribution does not depend on the position at which it was evaluated, ie. $[R]_{x=x_c}=[R]$.

@@

It is however straiforward, to setup these equations.

@@

Additionally bindings are generated such that the full power of the simulation tool can be accessed.

@@

\subsubsection{Homogenous and Inhomogenous Initial Cell Distributions}
To clarify what we mean by homogenous and inhomogenous initial configurations, we include snapshots of the simulation.

@@

        \includegraphics[width=\textwidth]{Figures-abm-homogenous-images-snapshot_00012000.png}
        \subcaption{Homogenous initial cell distribution, $D_R=D_I=30.0$\unit{\micro\metre\squared\per\minute}}

@@

        \includegraphics[width=\textwidth]{Figures-abm-homogenous-low-diffusion-images-snapshot_00012000.png}
        \subcaption{Homogenous initial cell distribution, $D_R=D_I=5.0$\unit{\micro\metre\squared\per\minute}}

>>>>

\setlength{\tabcolsep}{1.3pt}
\setlength{\columnsep}{0.4cm}
\usepackage{acronym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{array}
\usepackage{bm}
\usepackage{booktabs}
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off, font=small]{caption}
\usepackage{diagbox}
\usepackage{float}
\usepackage[T1]{fontenc}
% \usepackage[a4paper,inner=2cm,outer=2cm,top=2.5cm,bottom=2.5cm]{geometry}
\usepackage{graphicx}% Include figure files
\usepackage{here}
\usepackage{hyperref}
\usepackage{lipsum}
\usepackage{listings}
\usepackage{listings-rust}
\usepackage{mathtools}
\usepackage{mathrsfs}
\usepackage{multirow}
\usepackage{newfloat}
\usepackage{onimage}
\usepackage{siunitx}
\usepackage{subcaption}
\usepackage{tabularx}
\usepackage{tikz}
%\usepackage[dvipsnames]{xcolor}
\usepackage{xpatch}

@@

\newcommand{\beginsupplement}{%
  \renewcommand{\thesection}{}
  \setcounter{subsection}{0}
  \renewcommand{\thesubsection}{S\arabic{subsection}}
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}%
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}%
  \setcounter{listing}{0}
  \renewcommand{\thelisting}{S\arabic{listing}}
}
\newcommand{\mbx}{\mathbf{x}}
\newcommand{\mbm}{\mathbf{m}}
\newcommand{\mbp}{\mathbf{p}}
\newcommand{\mbu}{\mathbf{u}}
\newcommand{\mbF}{\mathbf{F}}
\newcommand{\mbH}{\mathbf{H}}
\newcommand{\cmt}[1]{\textcolor{red} {[#1]}}
\newcommand{\etal}{{\textit{et al. }}}

@@

\cortext[cor1]{Corresponding Author}
\fntext[fn1]{jonas.pleyer.org}
\affiliation[1]{
    organization={Freiburg Center for Data Analysis and Modeling},
    addressline={Ernst-Zermelo Str. 1},
    city={Freiburg im Breisgau},
    postcode={79104},
    country={Germany},
}

@@

\begin{abstract}
    Bacterial consortia are complex systems which are often comprised of multiple distinct species.
    Their dynamics on a population-based level often result from interactions that lead to
    interspecies cooperation or competition.
    Various empirical and inactivation models have been conceived and applied to problems related to
    food microbiology.
    However, the utilized models live on the population level, omitting any spatial effects such as
    heterogeneity or diffusion.
    In this work we present a novel generalized framework, called "pool model" which uses multiple
    distinct pools that can hold bacterial states of distinct species or extracellular nutrient
    abundance.
    It allows us to describe these systems mechanistically as variations of an underlying, more
    fundamental model.
    To illustrate the usefulness of the model, multiple different examples are reported.
    Finally, we construct an \acl{abm} for a given pool model, showcasing that the resulting model
    arises naturally from the individual behaviour of many cellular agents.
    We further apply this \ac{abm} to study the role of spatial effects on the dynamics of bacterial
    growth.
\end{abstract}

@@

To determine the system of equations which describe the system's dynamics, we consider the flows between the different states.

@@

As a result, species $A$ was able to suppress species $B$, consuming the larger amount of nutrients and reaching a larger bacterial count at the end.

@@

One of the examples of such mechanisms is the production of antimicrobial compounds, inhibitors or toxins~\cite{wloch-salamon_effect_2008, chao_structured_1981}.

@@

This mutual activation benefits all involved species and can be easily included in our mathematical framework.

@@

nutrients~\cite{htuson_bacteriasurface_2013}.

@@

\textit{B. subtilis} which are capable of chemotactic motion driven by oxygen

@@

The bacterial social interactions such as release of toxins or public goods can also

@@

% ------------------------------------------------------------------------------------------------
\subsection{Constructing an Agent-Based Pool Model}
The following sections will demonstrate how a given \ac{ode} system can be translated into a
custom-built \ac{abm} which has only a few more overall parameters.
This procedure is a blueprint which can be followed to construct \acp{abm} for similar
\ac{ode}-based systems.\\
To construct our \ac{abm}, we use \texttt{cellular\_raza}~\cite{Pleyer2025}.
It allows us to define our bacterial agents from a clean slate, without having to rely on and inherit
existing work.
With this flexibility, researchers are able to construct minimal models using only as many parameters which are necessary.
This ability, to precisely define and control the exact behavior of our cell-agents, enables the following parts.

@@

Additionally, we require our external components to be freely diffusible with constants $D_R,D_I$.

@@

In subfigure (c), due to the heterogeneous cell distribution (see also~\ref{fig:supplement-spatial-initial-config-cells}), overall inhibition of species $B$ is significantly reduced.

@@

This is expected, since processes such as diffusion and cells growing outward introduce another time component which is simply assumed to be instantaneous in the \ac{ode} model.

@@

        homogenous (Hom.), heterogeneous (Het.) or randomized (Rand.) initial configurations.

@@

The pool model presented here was used to describe the dynamics of 6 systems containing established models of bacterial growth.
Comparing our approach with the Baranyi-Roberts model which is widely used in food science, it is possible to obtain results which provide growth curves of very similar shape.
However, their analytical form is not completely identical.
The inclusion of additional metabolic processes or larger number of cell-species can be achieved by extending the list of possible states and defining transitions between them.
In addition, we have shown how to treat interspecies interactions in the form of cooperation and competition.
By constructing an \ac{abm}, we have exemplified how close the devised model is to the individual biological processes that occur within the bacteria.
This mechanistic single-cell level of understanding of our \ac{abm} is a perspective which can not be easily replicated by other spatial models such as \acp{pde}.
We have investigated the effect of spatial effects in the form of varying diffusion parameters and different initial spatial distributions of bacteria.
Our findings reinforce that the \ac{ode} model is a good simplification in the highly-diffusive regime within a mostly homogenous spatial bacterial distribution.
However, deviations to these assumptions can lead to non-negligible effects, influencing the growth of the colony.

@@

We introduced the pool model as a novel framework to unify existing approaches and allows researchers to follow a mechanistic approach during model development.
It is able to capture the effects of many established models and provides a lens which is useful to generalize existing systems and results in better overall explainability.
By modeling effects such as mutual inhibition on a single-cell level within an \ac{abm}, we have exemplified that the pool model directly builds on these more fundamental concepts.
It can thus aid in the mechanistic understanding of the underlying processes, allowing researchers to deepen their knowledge about microbial growth within food science.
Going forward, this framework can provide a starting point for a more abstract mathematical formulation which should be able to cover a much wider class of systems.

@@

This means that every contribution does not depend on the position at which it was evaluated, i.e. $[R]_{x=x_c}=[R]$.

@@

It is however straightforward, to set up these equations.

@@

Additionally, bindings are generated such that the full power of the simulation tool can be accessed.

@@

\subsubsection{Homogeneous and Inhomogeneous Initial Cell Distributions}
To clarify what we mean by homogeneous and inhomogeneous initial configurations, we include snapshots of the simulation.

@@

        \includegraphics[width=\textwidth]{Figures-abm-homogeneous-images-snapshot_00012000.png}
        \subcaption{Homogeneous initial cell distribution, $D_R=D_I=30.0$\unit{\micro\metre\squared\per\minute}}

@@

        \includegraphics[width=\textwidth]{Figures-abm-homogeneous-low-diffusion-images-snapshot_00012000.png}
        \subcaption{Homogeneous initial cell distribution, $D_R=D_I=5.0$\unit{\micro\metre\squared\per\minute}}
